SciPy 1.0---Fundamental Algorithms for Scientific Computing in Python

### 2019-05-04 scipy-1.0/paper.tex
change section title to Relation with downstream projects

Python, Cython, Fortran, C and C++ are the programming languages used to
implement scientific algorithms in the SciPy library. An analysis of our
codebase using the \texttt{linguist} library\cite{linguistref} provides a 
detailed breakdown as \% composition by programming language in 
SciPy (Figure~\ref{fig:linguist}).

\begin{figure}[H]
    \centering
		\includegraphics[width=\textwidth]{static/composition}
    \caption{The breakdown of programming languages used in the
             SciPy library determined using the linguist library.
    	 Small ($<0.5 \%$) amounts of TeX, Matlab, Shell,
    	 and Makefile are excluded for clarity and mostly
    	 provide supporting roles in tests, building, and
    	 documentation.}
    \label{fig:linguist}
\end{figure}

For implementing new functionality, we have a clear order of language
preference.  First Python, if performance is not an issue. If it is, then in
order of preference: Cython, C, C++, Fortran. The main motivation for this is
maintainability: Cython has the highest abstraction level and most Python
developers will understand it. C is also widely known, and easier to deal with
for the core development team than C++ and especially Fortran.

While it is not surprising that Python is heavily used in SciPy, the usage
distribution of the other languages warrants some discussion. Fortran is an
extremely well-established scientific programming language, both for historical
reasons and because of its continued excellent
performance\cite{Koelbel:1993:HPF:562354}. We wrap the FFTPACK Fortran library
for performing Fourier transforms\cite{SWARZTRAUBER198445, SWARZTRAUBER198251}
since this library has been a standard in the field for 33 years and has a
license that is compatible with our own. Likewise, we wrap the Fortran source
for ODEPACK\cite{citeulike:2644528} as it has been  trusted for the initial
value problem for ordinary differential equation systems for 30 years. For
similar reasons we also wrap the Fortran libraries
QUADPACK\cite{1983qspa.book.....P} (for numerical integration of
one-dimensional functions), FITPACK\cite{Dierckx:1993:CSF:151103}
(curve-fitting / interpolation), ODRPACK\cite{ODRPACK_Boggs} (orthogonal
distance regression), MINPACK\cite{osti_6997568} (minimization of linear and
nonlinear equations), ARPACK\cite{leh:sor:yan96} (solving large scale
eigenvalue problems), ALGORITHM 644\cite{Amos:1986:APP:7921.214331} (handling
Bessel Functions), and CDFLIB\cite{CDFLIB_site} (evaluation of cumulative
density functions).

Rounding out the top three languages in SciPy is C, which is also extremely
well-established over several decades\cite{Kernighan:1988:CPL:576122} of
scientific computing. Alongside its objected-oriented relative C++, C can be
leveraged in Python after being exposed using the Cython language. Cython has
been described as a creole language that mixes the best parts of Python and
lower-level C / C++ paradigms\cite{behnel2011cython}. We thus often use Cython
as a glue between well-established low-level scientific computing libraries and
the Python interface offered by SciPy. We also use Cython to enable performance
enhancements in Python code, especially for cases where heavily used inner
loops benefit from a compiled code with static typing. Some of the C libraries
that we wrap in SciPy include trlib\cite{doi:10.1080/10556788.2018.1449842}
(iterative solving of the trust region problem),
SuperLU\cite{li05,superlu_ug99} (solution of large, sparse, nonsymmetric
systems of linear equations), Qhull\cite{Barber:1996:QAC:235815.235821}
(computational geometry), and Cephes\cite{cephes_netlib} (mathematics
algorithms).

Therefore, the relative abundance of different programming languages in the
SciPy library results from a combination of the usage of powerful performance
enhancing languages that interact well with Python (i.e., Cython), and the
usage of languages (and their libraries) that have proven reliable and
performant over many decades. The position that SciPy occupies near the
foundation of the scientific Python ecosystem is such that adoption of new
languages or major dependencies is generally unlikely---our choices are strongly
driven by long-term stability. GPU acceleration, new transpiling libraries, and
the latest JIT compilation approaches (i.e.,
Numba\cite{Lam:2015:NLP:2833157.2833162}) are very powerful, but currently fall
outside the remit of the main SciPy library. That said, we have recently
increased our efforts to support compatibility with some of these options, and
having our full test suite pass with the PyPy JIT
compiler\cite{Bolz:2009:TMP:1565824.1565827} is now a requirement in our
development workflow.

>>>>

Figure~\ref{fig:linguist} provides a breakdown of the programming languages used
in SciPy. While it is not surprising that Python is predominant, the usage distribution 
of the other languages warrants discussion.

\begin{figure}[H]
    \centering
		\includegraphics[width=\textwidth]{static/composition}
    \caption{The breakdown of programming languages used in the
             SciPy library determined using the \texttt{linguist} library.
    	 Small ($<0.5 \%$) amounts of TeX, Matlab, Shell,
    	 and Makefile are excluded for clarity and mostly
    	 provide supporting roles in tests, building, and
    	 documentation.}
    \label{fig:linguist}
\end{figure}

Fortran, despite its age, is still a high-performance scientific programming language with 
continued contemporary usage\cite{Koelbel:1993:HPF:562354}. Thus, we wrap the following excellent, field-tested Fortran
libraries in order to provide Python convenience while benefiting from their performance: 
FFTPACK\cite{SWARZTRAUBER198445, SWARZTRAUBER198251}, % for performing Fourier transforms, 
ODEPACK\cite{citeulike:2644528}, % for solving ordinary differential equation initial value problems,
QUADPACK\cite{1983qspa.book.....P}, % for numerical integration of one-dimensional functions,
FITPACK\cite{Dierckx:1993:CSF:151103}, % for curve-fitting and interpolation,
ODRPACK\cite{ODRPACK_Boggs}, % for orthogonal distance regression, 
MINPACK\cite{osti_6997568}, % for minimization of linear and nonlinear equations, 
ARPACK\cite{leh:sor:yan96}, % for solving large scale eigenvalue problems,
ALGORITHM 644\cite{Amos:1986:APP:7921.214331}, and % for Bessel Functions, and 
CDFLIB\cite{CDFLIB_site}. % for evaluation of cumulative density functions.

Rounding out the top three languages in SciPy is C, which is also extremely
well-established over several decades\cite{Kernighan:1988:CPL:576122} of
scientific computing. The C libraries that we wrap in SciPy include 
trlib\cite{doi:10.1080/10556788.2018.1449842}, % (iterative solving of the trust region problem),
SuperLU\cite{li05,superlu_ug99}, % (solution of large, sparse, nonsymmetric systems of linear equations),
Qhull\cite{Barber:1996:QAC:235815.235821}, and % (computational geometry) 
Cephes\cite{cephes_netlib}. % (mathematics algorithms).

Cython has been described as a creole language that mixes the best parts of Python and
lower-level C / C++ paradigms\cite{behnel2011cython}. We often use Cython
as a glue between well-established low-level scientific computing libraries written
in C / C++ and the Python interface offered by SciPy. We also use Cython to enable performance
enhancements in Python code, especially for cases where heavily used inner
loops benefit from a compiled code with static typing.

For implementing new functionality, Python is the still the language of choice. If Python 
performance is an issue, then we prefer the use of Cython followed by C, C++, or Fortran (in that
order). The main motivation for this is maintainability: Cython has the 
highest abstraction level and most Python developers will understand it. C is 
also widely known, and easier for the current core development team to manage
than C++ and especially Fortran.

The distribution of secondary programming languages in SciPy is a compromise between
a powerful, performance-
enhancing language that interacts well with Python (i.e. Cython) and the
usage of languages (and their libraries) that have proven reliable and
performant over many decades. The position that SciPy occupies near the
foundation of the scientific Python ecosystem is such that adoption of new
languages or major dependencies is generally unlikely---our choices are strongly
driven by long-term stability. GPU acceleration, new transpiling libraries, and
the latest JIT compilation approaches (i.e.,
Numba\cite{Lam:2015:NLP:2833157.2833162}) are very powerful, but currently fall
outside the remit of the main SciPy library. That said, we have recently
increased our efforts to support compatibility with some of these options, and
having our full test suite pass with the PyPy JIT
compiler\cite{Bolz:2009:TMP:1565824.1565827} is now a requirement in our
development workflow.

### 2019-05-04 scipy-1.0/poly.tex

At the lowest level of the new design are classes which represent univariate
piecewise polynomials: \code{PPoly} (SciPy 0.13)\cite{scipy-gh5827},
\code{BPoly} (SciPy 0.13) and \code{BSpline} (SciPy 0.19)\cite{scipy-gh3174},
which allow
efficient vectorized evaluations, differentiation, integration and root-finding.
\code{PPoly} represents piecewise polynomials in the power basis in terms of
breakpoints and coefficients at each interval. \code{BPoly} is similar, and
represents piecewise polynomials in the Bernstein basis (which is suitable
for e.g., constructing Bezier curves). \code{BSpline} represents spline
curves, i.e., linear combinations of b-spline basis elements.\cite{deBoor1978}

>>>>

At the lowest level of the new design are classes which represent univariate
piecewise polynomials: \code{PPoly} (SciPy 0.13)\cite{scipy-gh2885},
\code{BPoly} (SciPy 0.13) and \code{BSpline} (SciPy 0.19)\cite{scipy-gh3174},
which allow
efficient vectorized evaluations, differentiation, integration and root-finding.
\code{PPoly} represents piecewise polynomials in the power basis in terms of
breakpoints and coefficients at each interval. \code{BPoly} is similar, and
represents piecewise polynomials in the Bernstein basis (which is suitable
for e.g., constructing Bezier curves). \code{BSpline} represents spline
curves, i.e., linear combinations of B-spline basis elements.\cite{deBoor1978}
